---
title: "Reproducible codes of the paper: Explainable expected goal models for performance analysis in football analytics"
author: "Mustafa Cavus"
date: "June 14, 2022"
output: html_document
---

# Necessary Packages

```{r include=FALSE}
library(tidyverse)      # data manipulation
library(ggplot2)        # data visualization
library(hrbrthemes)     # customization of plot theme
library(ROSE)           # over and under balancing data 
library(forester)       # training tree-based models (from GitHub: https://github.com/ModelOriented/forester)
library(DALEX)          # using XAI tools
library(ingredients)    # creating CP and AP 
library(worldfootballR) # scraping shot data  (from GitHub: https://github.com/JaseZiv/worldfootballR)
```


# Dataset

We focus in our paper on 315,430 shots-related event data (containing 33,656 goals  of total shots) from the 12,655 matches in 7 seasons between 2014-15 and 2020-21 from the top-five European football leagues which are Serie A, Bundesliga, La Liga, English Premier League, Ligue 1. The dataset is collected from Understat by using the R-package `worldfootballR` and excluded the 1,012 shots resulting in own goals due to their unrelated pattern from the concept of the model. The following function is used for scraping the data from the leagues over 7 seasons:

```{r eval=FALSE, include=FALSE}
# Ligue 1
ligue1_2020_shot_location <- understat_league_season_shots(league = "Ligue 1", season_start_year = 2020)
ligue1_2019_shot_location <- understat_league_season_shots(league = "Ligue 1", season_start_year = 2019)
ligue1_2018_shot_location <- understat_league_season_shots(league = "Ligue 1", season_start_year = 2018)
ligue1_2017_shot_location <- understat_league_season_shots(league = "Ligue 1", season_start_year = 2017)
ligue1_2016_shot_location <- understat_league_season_shots(league = "Ligue 1", season_start_year = 2016)
ligue1_2015_shot_location <- understat_league_season_shots(league = "Ligue 1", season_start_year = 2015)
ligue1_2014_shot_location <- understat_league_season_shots(league = "Ligue 1", season_start_year = 2014)

# Serie A
seriea_2020_shot_location <- understat_league_season_shots(league = "Serie A", season_start_year = 2020)
seriea_2019_shot_location <- understat_league_season_shots(league = "Serie A", season_start_year = 2019)
seriea_2018_shot_location <- understat_league_season_shots(league = "Serie A", season_start_year = 2018)
seriea_2017_shot_location <- understat_league_season_shots(league = "Serie A", season_start_year = 2017)
seriea_2016_shot_location <- understat_league_season_shots(league = "Serie A", season_start_year = 2016)
seriea_2015_shot_location <- understat_league_season_shots(league = "Serie A", season_start_year = 2015)
seriea_2014_shot_location <- understat_league_season_shots(league = "Serie A", season_start_year = 2014)

# Bundesliga
bundesliga_2020_shot_location <- understat_league_season_shots(league = "Bundesliga", season_start_year = 2020)
bundesliga_2019_shot_location <- understat_league_season_shots(league = "Bundesliga", season_start_year = 2019)
bundesliga_2018_shot_location <- understat_league_season_shots(league = "Bundesliga", season_start_year = 2018)
bundesliga_2017_shot_location <- understat_league_season_shots(league = "Bundesliga", season_start_year = 2017)
bundesliga_2016_shot_location <- understat_league_season_shots(league = "Bundesliga", season_start_year = 2016)
bundesliga_2015_shot_location <- understat_league_season_shots(league = "Bundesliga", season_start_year = 2015)
bundesliga_2014_shot_location <- understat_league_season_shots(league = "Bundesliga", season_start_year = 2014)

# La Liga
laliga_2020_shot_location <- understat_league_season_shots(league = "La liga", season_start_year = 2020)
laliga_2019_shot_location <- understat_league_season_shots(league = "La liga", season_start_year = 2019)
laliga_2018_shot_location <- understat_league_season_shots(league = "La liga", season_start_year = 2018)
laliga_2017_shot_location <- understat_league_season_shots(league = "La liga", season_start_year = 2017)
laliga_2016_shot_location <- understat_league_season_shots(league = "La liga", season_start_year = 2016)
laliga_2015_shot_location <- understat_league_season_shots(league = "La liga", season_start_year = 2015)
laliga_2014_shot_location <- understat_league_season_shots(league = "La liga", season_start_year = 2014)

# La Liga
epl_2020_shot_location <- understat_league_season_shots(league = "EPL", season_start_year = 2020)
epl_2019_shot_location <- understat_league_season_shots(league = "EPL", season_start_year = 2019)
epl_2018_shot_location <- understat_league_season_shots(league = "EPL", season_start_year = 2018)
epl_2017_shot_location <- understat_league_season_shots(league = "EPL", season_start_year = 2017)
epl_2016_shot_location <- understat_league_season_shots(league = "EPL", season_start_year = 2016)
epl_2015_shot_location <- understat_league_season_shots(league = "EPL", season_start_year = 2015)
epl_2014_shot_location <- understat_league_season_shots(league = "EPL", season_start_year = 2014)

# combining data
raw_data <- rbind(ligue1_2020_shot_location,
                  ligue1_2019_shot_location,
                  ligue1_2018_shot_location,
                  ligue1_2017_shot_location,
                  ligue1_2016_shot_location,
                  ligue1_2015_shot_location,
                  ligue1_2014_shot_location,
                  
                  seriea_2020_shot_location,
                  seriea_2019_shot_location,
                  seriea_2018_shot_location,
                  seriea_2017_shot_location,
                  seriea_2016_shot_location,
                  seriea_2015_shot_location,
                  seriea_2014_shot_location,
                  
                  bundesliga_2020_shot_location,
                  bundesliga_2019_shot_location,
                  bundesliga_2018_shot_location,
                  bundesliga_2017_shot_location,
                  bundesliga_2016_shot_location,
                  bundesliga_2015_shot_location,
                  bundesliga_2014_shot_location,
                  
                  laliga_2020_shot_location,
                  laliga_2019_shot_location,
                  laliga_2018_shot_location,
                  laliga_2017_shot_location,
                  laliga_2016_shot_location,
                  laliga_2015_shot_location,
                  laliga_2014_shot_location,
                  
                  epl_2020_shot_location,
                  epl_2019_shot_location,
                  epl_2018_shot_location,
                  epl_2017_shot_location,
                  epl_2016_shot_location,
                  epl_2015_shot_location,
                  epl_2014_shot_location)
```


# Pre-processing of the raw dataset

This section introduces the dataset and how it is pre-processed. First data is imported from a .csv file is `raw_data`, then the features `distanceToGoal` and `angleToGoal` are extracted from the coordinated `X` and `Y`. The features `status`, `distanceToGoal`, `angleToGoal`, `h_a`, `shotType`, `lastAction`, `minute`, `league`, and `season` are prepared for analysis and modeling.

```{r include=FALSE}
shot_stats <- dataset %>% filter(result != "OwnGoal") %>%
  mutate(status = ifelse(result == "Goal", 1, 0)) %>%
  mutate(distanceToGoal = sqrt((105 - (X * 105)) ^ 2 + (32.5 - (Y * 68)) ^ 2)) %>%
  mutate(angleToGoal = abs(atan((7.32 * (105 - (X * 105))) / ((105 - (X * 105))^2 + (32.5 - (Y * 68)) ^ 2 - (7.32 / 2) ^ 2)) * 180 / pi)) %>%
  mutate(h_a = factor(h_a),
         situation = factor(situation),
         shotType = factor(shotType),
         lastAction = factor(lastAction),
         minute = as.numeric(minute)) %>%
  select(status, minute, h_a, situation, shotType, lastAction, 
         distanceToGoal, angleToGoal, league, season, match_id, result, player_id)
```


# Preparing sets for model training

```{r}
# preparing train set of original dataset
train_data <- shot_stats %>%
   select(status, minute, h_a, situation, shotType, lastAction, 
          distanceToGoal, angleToGoal)

# preparing train set of under-sampled dataset
set.seed(123)
under_train_data <- ovun.sample(status ~ ., data = train_data, method = "under")

# preparing train set of over-sampled dataset
set.seed(123)
over_train_data <- ovun.sample(status ~ ., data = train_data, method = "over")
```


# Model training

## Modifications on {forester}

We changed and expanded some functions of the forester package. You can see the reasons for this action below:

* The forester returns the predicted labels, we changed this with predicted probabilities to calculate the performance metrics which are based on probabilities such as log-loss, Brier score and MCC.

* The forester returns only the output of the best performing model in terms of the value of intended metric, we expanded it to return the output of all models for comparing their performance with the additional metrics.

* After under-sample the dataset, the ranger changes the reference class in the model and causes a inconsistency. Thus, we add an argument to the make_ranger and forester functions to control the reference class.

```{r}
source("evaluate.R")
source("forester.R")
source("make_ranger.R")
source("make_xgboost.R")
source("make_lightgbm.R")
source("make_catboost.R")
source("model_performancex.R")
```


We use the forester `forester` AutoML tool to train various tree-based classification models from `XGBoost`, `randomForest`, `LightGBM`, and `CatBoost` libraries.

```{r}
# training tree-based models on original dataset
set.seed(123)
original_model <- forester(data   = train_data,
                           target = "status",
                           type   = "classification")

# training tree-based models on under-sampled dataset
set.seed(123)
under_model <- forester(data    = under_train_data$data,
                        target  = "status",
                        type    = "classification",
                        refclass = "")

# training tree-based models on over-sampled dataset
set.seed(123)
over_model <- forester(data   = over_train_data$data,
                       target = "status",
                       type   = "classification")
```


# Table II: The summary statistics of the shots and goals, such as the number of matches, shots, goals, the mean of shots and goals per match and the conversion percent of a shot to goal for per league over seven seasons

```{r}
dataset %>%
  filter(result != "OwnGoal") %>%
  group_by(league) %>%
  summarise(nmatch = length(unique(match_id)), # number of matches
            nshot  = length(match_id),         # number of shots
            mshot  = nshot / nmatch,           # mean of shots per match
            ngoal  = sum(result == "Goal"),    # number of goals
            mgoal  = ngoal / nmatch,           # mean of goals per match
            conv   = ngoal / nshot * 100)      # convertion percent of the shots to goal
```


# Table III: Performance of trained xG models

```{r}
# performance of random forest model
# on over-sampled data
model_performancex(over_model$model3)
```

```{r}
# on under-sampled data
model_performancex(under_model$model3)
```

```{r}
# on original data
model_performancex(original_model$model3)
```


```{r}
# performance of catboost model
# on over-sampled data
model_performancex(over_model$model1)
```

```{r}
# on under-sampled data
model_performancex(under_model$model1)
```

```{r}
# on original data
model_performancex(original_model$model1)
```


```{r}
# performance of xgboost model
# on over-sampled data
model_performancex(over_model$model2)
```

```{r}
# on under-sampled data
model_performancex(under_model$model2)
```

```{r}
# on original data
model_performancex(original_model$model2)
```


```{r}
# performance of lightgbm model
# on over-sampled data
model_performancex(over_model$model4)
```

```{r}
# on under-sampled data
model_performancex(under_model$model4)
```

```{r}
# on original data
model_performancex(original_model$model4)
```


# Table IV: Performance Comparison of the proposed xG model with the models in the literature

```{r}
# calculation of MAE values of the proposed models
# random forest
median(abs(predict(over_model$model3, 
                   over_model$test_data) - over_model$test_data$status))

# catboost
median(abs(predict(over_model$model1, 
                   over_model$test_data) - over_model$test_data$status))

# xgboost
median(abs(predict(over_model$model2, 
                   over_model$test_data[,-1]) - over_model$test_data$status))

# lightgbm
median(abs(predict(over_model$model4, 
                   over_model$test_data[,-1]) - over_model$test_data$status))
```


# Table V: The End-of-match statistics of Schalke 04 vs. Bayern Munich in the match is played on Jan 24, 2021

```{r}
# match_id = 15298: 01.24.2021 - Schalke 04 - Bayern Munich
# stats of Schalke 04
shot_stats %>% 
  filter(match_id == 15298, h_a == "h") %>% 
  summarise(ngoal = sum(result == "Goal"),
            nshot = length(match_id),
            matg  = mean(angleToGoal),
            mdtg  = mean(distanceToGoal))
```

```{r}
# stats of Bayern Munich
shot_stats %>% 
  filter(match_id == 15298, h_a == "a") %>% 
  summarise(ngoal = sum(result == "Goal"),
            nshot = length(match_id),
            matg  = mean(angleToGoal),
            mdtg  = mean(distanceToGoal))
```


# Table VI: The end-of-season statistics of Burak Yilmaz (BY), Lionel Messi (LM), and Robert Lewandowski (RL) in the season of 2020-21

```{r}
# player_id == 8637 -> Burak Yilmaz
shot_stats %>%
  filter(player_id == 8637, season == 2020) %>%
  summarise(ngame = length(unique(match_id)),
            ngoal = sum(result == "Goal"),
            nshot = length(match_id),
            matg  = mean(angleToGoal),
            mdtg  = mean(distanceToGoal))
```

```{r}
# player_id == 2097 -> Lionel Messi
shot_stats %>%
  filter(player_id == 2097, season == 2020) %>%
  summarise(ngame = length(unique(match_id)),
            ngoal = sum(result == "Goal"),
            nshot = length(match_id),
            matg  = mean(angleToGoal),
            mdtg  = mean(distanceToGoal))
```

```{r}
# player_id == 227 -> Robert Lewandowski
shot_stats %>%
  filter(player_id == 227, season == 2020) %>%
  summarise(ngame = length(unique(match_id)),
            ngoal = sum(result == "Goal"),
            nshot = length(match_id),
            matg  = mean(angleToGoal),
            mdtg  = mean(distanceToGoal))
```


# Figure 1: The distribution of angle to goal and distance to goal of shots regarding goal status in the last seven seasons of top-five European football leagues

```{r}
shot_vis <- data.frame(sta = as.factor(rep(shot_stats$status, 2)),
                       obs = c(shot_stats$distanceToGoal, 
                               shot_stats$angleToGoal),
                       vty = as.factor(c(rep("Distance to goal", length(shot_stats$distanceToGoal)),
                              rep("Angle to goal", length(shot_stats$angleToGoal)))),
                       lea = rep(shot_stats$league, 2),
                       sea = rep(shot_stats$season, 2))

shot_vis$lea[shot_vis$lea == "La_liga"] <- "La Liga"
shot_vis$lea[shot_vis$lea == "Ligue_1"] <- "Ligue 1"
shot_vis$lea[shot_vis$lea == "Serie_A"] <- "Serie A"

ggplot(shot_vis, 
       aes(x = obs, 
           group = sta, 
           fill = sta)) + 
  geom_density(alpha = 0.5) + 
  theme_bw() + 
  scale_fill_brewer(palette = "Set1",
                    name = "Goal Status", 
                    labels = c("No goal", "Goal")) + 
  theme(legend.position = "bottom") +
  labs(x = "Value (meter or angle)",
       y = "Density") + 
  facet_grid(lea ~ vty, 
             scales="free_x") 
```


# Figure 3: The aggregated xG profiles of Schalke 04 and Bayern Munich for angle and distance to goal in the match is played on Jan 24, 2021

```{r}
# subsetting the shots of Schalke 04 and Bayern Munich
selected_observations_schalke <- shot_stats %>%
  filter(h_a == "h", match_id == 15298)
selected_observations_bayern <- shot_stats %>%
  filter(h_a == "a", match_id == 15298)

# creating ceteris-paribus profiles for the shots of Schalke 04 and Bayern Munich
cp_schalke <- ingredients::ceteris_paribus(over_model$model3,
                                           selected_observations_schalke)
cp_bayern  <- ingredients::ceteris_paribus(over_model$model3,
                                           selected_observations_bayern)

# re-labeling the profiles
cp_schalke$`_label_` <- "Schalke 04"
cp_bayern$`_label_`  <- "Bayern Munich"

# creating aggregated profiles for the shots of Schalke 04 and Bayern Munich
# for "distanceToGoal" feature
ap_dtg_schalke <- ingredients::aggregate_profiles(cp_schalke,
                                                  variables = "distanceToGoal",
                                                  type = "partial")
ap_dtg_bayern  <- ingredients::aggregate_profiles(cp_bayern,
                                                  variables = "distanceToGoal",
                                                  type = "partial")

# drawing plot of APs for "distanceToGoal" feature
plot(ap_dtg_schalke, ap_dtg_bayern)
```

```{r}
# creating aggregated profiles for the shots of Schalke 04 and Bayern Munich
# for "angleToGoal" feature
ap_atg_schalke <- ingredients::aggregate_profiles(cp_schalke,
                                                  variables = "angleToGoal",
                                                  type = "partial")
ap_atg_bayern  <- ingredients::aggregate_profiles(cp_bayern,
                                                  variables = "angleToGoal",
                                                  type = "partial")

# drawing the comparison plot of APs for "angleToGoal" feature
plot(ap_atg_schalke, ap_atg_bayern)
```


# Figure 4: The End-of-season statistics of Burak Yilmaz (BY), Lionel Messi (LM), and Robert Lewandowski (RL) in the season of 2020-21

```{r}
# subsetting the shots of Burak Yilmaz (by), Lionel Messi (lm) 
# and Robert Lewandowski (rl) in the season of 2020-21
selected_observations_by <- shot_stats %>%
  filter(player_id == 8637, season == 2020)
selected_observations_lm <- shot_stats %>%
  filter(player_id == 2097, season == 2020)
selected_observations_rl <- shot_stats %>%
  filter(player_id == 227, season == 2020)

# creating ceteris-paribus profiles for the shots of 
# Burak Yilmaz, Lionel Messi, and Robert Lewandowski
cp_by <- ceteris_paribus(over_model$model3,
                         selected_observations_by)
cp_lm <- ceteris_paribus(over_model$model3,
                         selected_observations_lm)
cp_rl <- ceteris_paribus(over_model$model3,
                         selected_observations_rl)

# re-labeling the profiles
cp_by$`_label_` <- "Burak Yilmaz"
cp_lm$`_label_` <- "Lionel Messi"
cp_rl$`_label_` <- "Robert Lewandowski"

# creating aggregated profiles for the shots of 
# Burak Yilmaz, Lionel Messi, and Robert Lewandowski
# for "distanceToGoal" feature
ap_dtg_by <- aggregate_profiles(cp_by,
                                variables = "distanceToGoal",
                                type = "partial")
ap_dtg_lm <- aggregate_profiles(cp_lm,
                                variables = "distanceToGoal",
                                type = "partial")
ap_dtg_rl <- aggregate_profiles(cp_rl,
                                variables = "distanceToGoal",
                                type = "partial")

# drawing the comparison plot of APs for "distanceToGoal" feature
plot(ap_dtg_by, ap_dtg_lm, ap_dtg_rl)
```

```{r}
# creating aggregated profiles for the shots of 
# Burak Yilmaz, Lionel Messi, and Robert Lewandowski
# for "angleToGoal" feature
ap_atg_by <- aggregate_profiles(cp_by,
                                variables = "angleToGoal",
                                type = "partial")
ap_atg_lm <- aggregate_profiles(cp_lm,
                                variables = "angleToGoal",
                                type = "partial")
ap_atg_rl <- aggregate_profiles(cp_rl,
                                variables = "angleToGoal",
                                type = "partial")

# drawing the comparison plot of APs for "angleToGoal" feature
plot(ap_atg_by, ap_atg_lm, ap_atg_rl)
```


# Figure 5: The behavior comparison of the random forest models trained on original, over-sampled, and under-sampled data in terms of PDP curves

```{r}
# creating an explainer for the random forest model on the original dataset
org_rf_explainer <- explain(model = original_model$model3,
                            data = original_model$test_data[,-1],
                            y = original_model$test_data$status,
                            type = "classification",
                            label = "original data")

# creating an explainer for the random forest model on the under-sampled dataset
under_rf_explainer <- explain(model = under_model$model3,
                              data = under_model$test_data[,-1],
                              y = under_model$test_data$status,
                              type = "classification",
                              label = "under-sampled")

# creating an explainer for the random forest model on the over-sampled dataset
over_rf_explainer <- explain(model = over_model$model3,
                             data = over_model$test_data[,-1],
                             y = over_model$test_data$status,
                             type = "classification",
                             label = "over-sampled")

# drawing the comparison plot of PDPs for "angleToGoal" feature
  plot(model_profile(org_rf_explainer, variables = "angleToGoal"),
     model_profile(under_rf_explainer, variables = "angleToGoal"),
     model_profile(over_rf_explainer,  variables = "angleToGoal"))

# drawing the comparison plot of PDPs for "distanceToGoal" feature
plot(DALEX::model_profile(org_rf_explainer,   variables = "distanceToGoal"),
     DALEX::model_profile(under_rf_explainer, variables = "distanceToGoal"),
     DALEX::model_profile(over_rf_explainer,  variables = "distanceToGoal"))

# drawing the comparison plot of PDPs for "minute" feature
plot(DALEX::model_profile(org_rf_explainer,   variables = "minute"),
     DALEX::model_profile(under_rf_explainer, variables = "minute"),
     DALEX::model_profile(over_rf_explainer,  variables = "minute"))
```

